Cooperative Kernels: GPU Multitasking for Blocking Algorithms

### 2017-07-06 write-ups/FSE/paper.tex
Finished pass over section 5.

Instead of multitasking multiple separate kernels, we merge a set of
kernels into a megakernel---a single, monolithic kernel.  The
megakernel is launched with as many workgroups as can be occupant
concurrently.  One workgroup takes the role of the
scheduler\footnote{\TSAdded{We note that the scheduler requirements
    given in \mysec{\ref{sec:cooperativekernels}} are agnostic to
    whether the scheduling logic takes place on the CPU or GPU. To
    avoid expensive communication between GPU and host, we choose to
    implement the scheduler on the GPU.}  }, and the scheduling logic
is embedded as part of the megakernel.  The remaining workgroups act
as a pool of workers.  A worker repeatedly queries the scheduler to be
assigned a task.  A task corresponds to executing a cooperative or
non-cooperative kernel.  In the non-cooperative case, the workgroup
executes the relevant kernel function uninterrupted, then awaits
further work.  In the cooperative case, the workgroup either starts
from the kernel entry point or immediately jumps to a designated point
within the kernel, depending on whether the workgroup is an initial
workgroup of the kernel, or a forked workgroup.  In the latter case,
the new workgroup also receives a struct containing the values of all
relevant $\transmit$-annotated variables.

\myparagraph{Simplifying Assumptions}
%
For ease of implementation, our prototype supports multitasking a
single cooperative kernel with a single non-cooperative kernel (though
the non-cooperative kernel can be invoked many times).
%We assume all
%variables declared at the root lexical scope of a cooperative kernel
%are implicitly $\transmit$-annotated.
%this allows our
%implementation to piggy-back on top of Clang (see below) without
%requiring front-end changes.
We require that $\offerkill$, $\offerfork$ and
$\resizingglobalbarrier$ are called from the entry function of a
cooperative kernel.  This allows us to use $\keyword{goto}$ and
$\keyword{return}$ to direct threads into and out of the kernel.  With
these restrictions we can experiment with interesting irregular
algorithms (see \mysec\ref{sec:experiments}).  A non-mock
implementation of cooperative kernels would not use the megakernel
approach, so we do not deem the engineering effort associated with
lifting these restrictions to be worthwhile.

@@

To enable multitasking through cooperative kernels, the runtime (in
our case, the megakernel) must track the state of workgroups,
i.e.\ whether a workgroup is waiting or computing a kernel, maintain
consistent context states for each kernel, e.g.\ tracking the number
of active workgroups, and provide a safe way for these states to be
modified in response to $\offerfork$/$\offerkill$. We discuss these
issues, and describe the implementation of an efficient resizing
barrier. We describe how the scheduler would handle arbitrary
combinations of kernels, though as noted above our current
implementation is restricted to the case of two kernels.

@@

Incorporating $\offerfork$ and $\offerkill$ into such a barrier is
straight forward. Upon entering the barrier, the slaves first execute
$\offerkill$, possibly exiting. The master then waits for $M$ slaves
(the number of active workgroups), which may decrease due to
$\offerkill$ calls by the slaves, but will not increase. Once the
master observes that $M$ slaves have arrived, it knows that all other
workgroups are waiting to be released. The master executes
$\offerfork$, and the statement immediately following this
$\offerfork$ is a conditional that forces newly spawned workgroups to
join the slaves in waiting to be released. Finally, the master
releases all the slaves: the original slaves and the new slaves that
joined at $\offerfork$.

>>>>

Instead of multitasking multiple separate kernels, we merge a set of
kernels into a megakernel---a single, monolithic kernel.  The
megakernel is launched with as many workgroups as can be occupant
concurrently.  One workgroup takes the role of the
scheduler,\footnote{\TSAdded{We note that the scheduler requirements
    given in \mysec{\ref{sec:cooperativekernels}} are agnostic to
    whether the scheduling logic takes place on the CPU or GPU. To
    avoid expensive communication between GPU and host, we choose to
    implement the scheduler on the GPU.}  } and the scheduling logic
is embedded as part of the megakernel.  The remaining workgroups act
as a pool of workers.  A worker repeatedly queries the scheduler to be
assigned a task.  A task corresponds to executing a cooperative or
non-cooperative kernel.  In the non-cooperative case, the workgroup
executes the relevant kernel function uninterrupted, then awaits
further work.  In the cooperative case, the workgroup either starts
from the kernel entry point or immediately jumps to a designated point
within the kernel, depending on whether the workgroup is an initial
workgroup of the kernel, or a forked workgroup.  In the latter case,
the new workgroup also receives a struct containing the values of all
relevant $\transmit$-annotated variables.

\myparagraph{Simplifying Assumptions}
%
For ease of implementation, our prototype supports multitasking a
single cooperative kernel with a single non-cooperative kernel (though
the non-cooperative kernel can be invoked many times).
%We assume all
%variables declared at the root lexical scope of a cooperative kernel
%are implicitly $\transmit$-annotated.
%this allows our
%implementation to piggy-back on top of Clang (see below) without
%requiring front-end changes.
We require that $\offerkill$, $\offerfork$ and
$\resizingglobalbarrier$ are called from the entry function of a
cooperative kernel.  This allows us to use $\keyword{goto}$ and
$\keyword{return}$ to direct threads into and out of the kernel.  With
these restrictions we can experiment with interesting irregular
algorithms (see \mysec\ref{sec:experiments}).  A non-mock
implementation of cooperative kernels would not use the megakernel
approach, so we did not deem the engineering effort associated with
lifting these restrictions in our prototype to be worthwhile.

@@

To enable multitasking through cooperative kernels, the runtime (in
our case, the megakernel) must track the state of workgroups,
i.e.\ whether a workgroup is waiting or computing a kernel; maintain
consistent context states for each kernel, e.g.\ tracking the number
of active workgroups; and provide a safe way for these states to be
modified in response to $\offerfork$/$\offerkill$. We discuss these
issues, and describe the implementation of an efficient resizing
barrier. We describe how the scheduler would handle arbitrary
combinations of kernels, though as noted above our current
implementation is restricted to the case of two kernels.

@@

Incorporating $\offerfork$ and $\offerkill$ into such a barrier is
straightforward. Upon entering the barrier, the slaves first execute
$\offerkill$, possibly exiting. The master then waits for $M$ slaves
(the number of active workgroups), which may decrease due to
$\offerkill$ calls by the slaves, but will not increase. Once the
master observes that $M$ slaves have arrived, it knows that all other
workgroups are waiting to be released. The master executes
$\offerfork$, and the statement immediately following this
$\offerfork$ is a conditional that forces newly spawned workgroups to
join the slaves in waiting to be released. Finally, the master
releases all the slaves: the original slaves and the new slaves that
joined at $\offerfork$.
